Asymmetric relationships and their effects on coexistence

### 2023-05-07 paper/Paper.Rnw
Dan finished going through the paper, about to try cutting 550 words

\author[a]{Pimsupa Jasmin Albert}
\author[a,b,*]{Daniel C. Reuman}

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

@@

\keywords{asymmetric tail association, biodiversity, coexistence, competition, competitive exclusion, modern coexistence theory, niche differentiation, storage effects}

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships - between organisms and between organisms and their environments. 
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships and to understand how two variables interact, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what we have previously referred to as 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are extremely common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_AER} show that ATAs are common in ecology
and can influence Taylor's law and Moran effects; \cite{Ghosh_2020_Eco} argue that extinction risks can become elevated 
when metapopulations are subject to ATAs; and \cite{Ghosh_2021} show that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}g,h show contrasting examples of ATAs in nature using 
plankton population density time series. To our knowledge, the implications of ATAs for competition and coexistence have
not yet been investigated. Doing so is the overarching goal of this study.

@@

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
Notation for the whole paper is summarized in Table \ref{tab:notation}.
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}d-f), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}d-f).

@@

%***DAN: Dan made it to here on 2022 04 26

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
strictly about covariation. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT that was goal G1 of the Introduction.

@@

%the beta fucundities version
The \emph{beta fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$. The parameter $\eta_i$ represents the upper
bound on what fecundities could occur for species $i$.

@@

To achieve goal G3 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we used a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decompose
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

@@

%One-parameters plots
To display results for goal G3, we started by plotting the terms of 
(\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We again labelled regions of parameters space corresponding to
ATA rescue or exclusion.

@@

%Summary-type paragraph of main results
We produced and applied a new extension of modern coexistence theory (MCT) that 
quantifies the impact of asymmetric tail associations (ATAs) on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to a laboratory diatom system demonstrated that  
ATA contributions to storage effects and to species coexistence 
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, generally, including both ATA effects and
correlation \emph{per se}. The prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know can be important. ATAs contributions to coexistence were less important than other mechanisms 
in some contexts (the log-normal fecundities lottery model), but were very important in other contexts
(the beta fecundities lottery model), including our empirically
based example (the diatom system). Though future work should seek to understand precisely when ATAs are or are not important
for coexistence, our work demonstrates the overall importance of this new mechanism.

@@

%***DAN: Jasmin, please add to the list anyone you talked to at UO about this, and anyone who read it and commented 
%for you. Be generous here - even short convos merit an acknowlegement.
The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, Laurn Hallett, and 
Robin Snyder for helpful discussions. 
The authors especially thank Steve Ellner for key advice at several watershed moments in the development of the project. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, the California Department of Fish and
Wildlife Delta Science Program, and the Alexander von Humboldt Foundation.

@@

%***DAN: Jasmin, could you, with reasonable effort, move panels around? Specifically, put the panels which
%are now g ang h in the middle instead of on the right (and they will then be called panels d and e) and 
%put what are now panels d-f on the right (and thwill then be called f-h)? The reason is, we refer to
%g-h before we refer to d-f, and the reader may become confused by this, including possibly havig trouble 
%finding the parts of the caption that refer to g-h while getting confused about the parts of the caption 
%that refer to d-e, which are about something that has not even been introduced at the point at which the 
%reader first is referred th panels g-h. 
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (g,h)]. 
The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
pictured and then transforming the marginals as specified in Methods.
(d-f) Lottery model population simulations using $(B_1,B_2)$ obtained by transforming the 
respectively corresponding variables (a-c; see Introduction).
(g,h) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs differed by location, with (g) showing 
left-tail association and (h) showing right-tail association. \label{fig:pedagog1}}
\end{figure}

%***DAN: This looks pretty good except the histograms for some marginals do not really look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%***DAN: I replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

>>>>

\author[a,b]{Pimsupa Jasmin Albert}
\author[a,c,*]{Daniel C. Reuman}

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Environmental Studies Program and Department of Biology, Universit of Oregon}
\affil[c]{Laboratory of Populations, Rockefeller University}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

\noindent \textbf{Running head:} Asymmetric relationships and coexistence \\
\noindent \textbf{Key words:} asymmetric tail association, biodiversity, coexistence, competition, copula, modern coexistence theory, niche differentiation, storage effects \\
\noindent \textbf{Type of article:} Letter\\
\noindent \textbf{Words in Abstract:} 150; \textbf{Words in main text:} 5518; \textbf{Words in text boxes:} 0\\
\noindent \textbf{Number of references:} 39; \textbf{Number of figures:} 5; \textbf{Number of tables:} 1; \textbf{Number of text boxes:} 0\\
\noindent \textbf{Corresponding author:}
Daniel C. Reuman,
Department of Ecology and Evolutionary Biology,
University of Kansas,
2101 Constant Ave,
Lawrence, KS 66047;
Phone: 626-560-7084 \\
\noindent \textbf{Author emails:} pimsupaa@uoregon.edu, reuman@ku.edu\\
\noindent \textbf{Statement of Authorship:} The authors jointly carried out all parts of the study.\\
\noindent \textbf{Data accessibility statement:} All codes associated with this study will be released
in a Zenodo repository prior to publication.

\doublespacing

@@

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships - between organisms and between organisms and their environments. 
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships and to understand how two variables interact, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what we have previously referred to as 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are extremely common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_AER} show that ATAs are common in ecology
and can influence Taylor's law and Moran effects; \cite{Ghosh_2020_Eco} argue that extinction risks can become elevated 
when metapopulations are subject to ATAs; and \cite{Ghosh_2021} show that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series. To our knowledge, the implications of ATAs for competition and coexistence have
not yet been investigated. Doing so is the overarching goal of this study.

@@

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out 
simple simulations using a two-species model, the well-known lottery model. The simulations described in this
paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; the
precise nature of that influence is the subject of the rest of the paper.
The lottery model, which was originally introduced as a model of coral reef fish competing for a set of $N$ breeding territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
Notation for the whole paper is summarized in Table \ref{tab:notation}.
We assume for simplicity that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions for these fecundities were used in all 
simulations (see Methods
for details). Thus, the model setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).

@@

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
strictly about covariation. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}, 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation 
in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT that was goal G1 of the Introduction.

@@

%the beta fucundities version
The \emph{beta fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-0.5}(1-x)^{-0.5}$ on
the unit interval). So the beta fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$. The parameter $\eta_i$ represents the upper
bound on what fecundities could occur for species $i$.

@@

To achieve goal G3 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we used a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is an extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decompose
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

@@

%One-parameters plots
To display results for goal G3, we started by plotting the terms of 
(\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We again labelled regions of parameter space corresponding to
ATA rescue or exclusion.

@@

%Summary-type paragraph of main results
We produced and applied a new extension of modern coexistence theory (MCT) that 
quantifies the impact of asymmetric tail associations (ATAs) on species coexistence. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to a laboratory diatom system demonstrated that  
ATA contributions to storage effects and to species coexistence 
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, generally, including both ATA effects and
correlation \emph{per se}. The prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know can be important. ATAs contributions to coexistence were less important than other mechanisms 
in some contexts (the log-normal fecundities lottery model), but were very important in other contexts
(the beta fecundities lottery model), including our empirically
based example (the diatom system). Though future work should seek to understand precisely when ATAs are or are not important
for coexistence, our work demonstrates the overall importance of this new mechanism.

@@

The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, Laurn Hallett, and 
Robin Snyder for helpful discussions. 
The authors especially thank Steve Ellner for key advice at several watershed moments in the development of the project. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, the California Department of Fish and
Wildlife Delta Science Program, and the Alexander von Humboldt Foundation.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (d,e)]. 
The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
pictured and then transforming the marginals as specified in Methods.
(d,e) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs differed by location, with (d) showing 
left-tail association and (e) showing right-tail association.
(f-h) Lottery model population simulations using $(B_1,B_2)$ obtained by transforming the 
respectively corresponding variables (a-c; see Introduction).
\label{fig:pedagog1}}
\end{figure}

%***DAN: This looks pretty good except one of the histograms on c does not look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
%***DAN: Jasmin, also, we use \varphi in the text, but \phi here. These need to be made consistent, since they are different to
%a math person. The easiest way will be to change it in the figure, so I hope that can straightforwardly be done.
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%Dan replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}
